Modelling BCG vaccination in the UK

### 2019-12-04 rmds/07-direct-effect-bcg-change.Rmd
wordcount
direct effects corrections excluding major

## Read formated results
## See https://github.com/seabbs/DirectEffBCGPolicyChange/blob/master/vignettes/paper.Rmd
## for code required to reproduce 
## Chapter is an edited version of the orginal reproducible document.
##Wordcount: 4504
read_formated <- function(name) {
  path <- file.path("chapters/direct-eff-bcg-change", paste0(name, ".rds"))

@@

I combined notification data from the Enhanced TB Surveillance (ETS) system, with demographic data from the Labour Force Survey (LFS) to construct retrospective cohorts of individuals in England relevant to both the universal, and targeted vaccination programmes between Jan 1, 2000 and Dec 31, 2010. For each cohort, I estimated incidence rates over a 5 year follow-up period and used Poisson and Negative Binomial regression models in order to estimate the impact of the change in policy on TB. This work was adapted from a preprint^[Preprint: https://doi.org/10.1101/567511] supervised by Hannah Christensen and Ellen Brooks-Pollock. Nicky Welton provided guidance on the statistical methods used.

@@

### Constructing Retrospective cohorts

I constructed retrospective cohorts of TB cases and individuals using the ETS and the LFS. TB cases were extracted from the ETS based on date of birth and date of TB notification.

Cohort 1: individuals aged 14 between 2000 and 2004, who were notified with TB whilst aged between 14 and 19 years old.

Comparison cohort 1: individuals aged 14 between 2005 and 2010, who were notified with TB whilst aged between 14 and 19 years old.

Cohort 2: individuals born between 2005 and 2010, who were notified with TB whilst aged 0-5 years old

Comparison cohort 2: individuals born between 2000 and 2004, who were notified with TB whilst aged 0-5 years old

Each cohort was stratified by UK birth status, with both non-UK born and UK born cases assumed to have been exposed to England’s vaccination policy. Corresponding population cohorts were calculated using the LFS population estimates, resulting in 8 population level cohorts, each with 5 years of follow up (Table \@ref(tab:07-cohorts)).

@@

I calculated Incidence Rate Ratios (IRRs) for the change in incidence rates associated with the change in BCG vaccination policy (modelled as a binary breakpoint at the start of 2005) for both the UK born and non-UK born populations that were relevant to the universal programme, and for the targeted programme using a series of increasingly complex models. I considered the following covariates: age,[@PHE2016a; @Zwerling2011] incidence rates in both the UK born and non-UK born who were not in the age group of interest,[@PHE2016a] and year of study entry (as a random intercept). I first investigated a univariable Poisson model, followed by combinations of covariates (Table \@ref(tab:07-model-spec)). I also investigated a Negative Binomial model adjusting for the same covariates as in the best fitting Poisson model. The models were estimated with a Bayesian approach using Markov Chain Monte Carlo (MCMC), with default weakly informative priors (see  Section \@ref(prior-choice)). Model fit, penalised by model complexity, was assessed using the leave one out cross validation information criterion (LOOIC) and its standard error.[@Vehtari2016] Models were ranked by goodness of fit, using their LOOIC, with a smaller LOOIC indicating a better fit to the data after adjusting for the complexity of the model. No formal threshold for a change in the LOOIC was used, with changes in the LOOIC being evaluated in the context of their standard error. The inclusion of the change in policy in the best fitting model was tested by refitting the model excluding the change in policy and estimating the improvment in the LOOIC. Once the best fitting model had been identified I estimated the number of cases prevented, from 2005 until 2015, for each vaccination programme in the study population relevant to that programme (see Section \@ref(magnitude-estimation)).

@@

R `3.5.0` was used for all analysis.[@R] Reproducibility was ensured by using `R` package infrastructure^[Code: https://github.com/seabbs/DirectEffBCGPolicyChange]. Missing data imputation using a GBM was implemented using the `h2o` package (see Section \@ref(imp-uk-birth-status)).[@h2o2018] Incidence rates, with 95% confidence intervals, were calculated using the `epiR` package (see Chapter \@ref(data)).[@EpiR] The `brms` package,[@Burkner] and STAN,[@StanDevelopmentTeam2016] was used to perform MCMC. Models were run until convergence (4 chains with a burn in of 10,000, and 10,000 sampled iterations each), with convergence being assessed using trace plots and the R hat diagnostic.[@StanDevelopmentTeam2016] All numeric confounders were centered and scaled by their standard deviation, and age was adjusted for using single year of age categories.

@@

As I was imputing a single variable I reformulated the imputation as a categorical prediction problem. This allowed the use of more complex, high-performing models compared to those usually used for imputation, whilst also allowing the results to be validated using predictive modelling performance metrics. I included year of notification, sex, age, PHE Centre (PHEC), occupation, ethnic group, Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank), and risk factor count (risk factors considered; drug use, homelessness, alcohol misuse/abuse and prison). However, I could not account for a possible missing not at random mechanism not captured by these covariates. To train the model I first split the data with complete UK birth status into a training set (80%), a calibration set (5%), and a test set (15%). I then fit a gradient boosted machine with 10,000 trees, early stopping (at a precision of $1e-5$, with 10 stopping rounds), a learning rate of 0.1, and a learn rate annealing of 0.99. Gradient boosted machines are a tree based method that can incorporate complex non-linear relationships and interactions. Much like a random forest model they work by ensembling a group of trees, but unlike a random forest model each tree is additive aiming to reduce the residual loss from previous trees. Once the model had been fit to the training set I performed platt scaling using the calibration dataset. Our fitted imputation model had a Logloss of 0.28 on the test set, with an AUC of 0.93, both of which indicate a robust out of bag performance. I found that ethnic group was the most important variable for predicting UK birth status, followed by age and PHEC.

@@

I estimated the magnitude of the estimated impact from the change in BCG policy by applying the IRR estimates from the best fitting model for each cohort to the observed number of notifications from 2005 until 2015 in the study population. For the cohorts relevant to the universal school-age vaccination scheme I estimated the number of prevented cases by first aggregating cases ($C_O$) and then using the following equation,

\[ C^i_P = C_O * (1 - I^i),\ \text{Where}\ i = e,\ l,\ u.\]

@@

In the UK born cohort relevant to universal vaccination there was some evidence, across all models that adjusted for age, that ending the scheme was associated with a modest increase in TB rates (Table \@ref(tab:07-summary-universal-ukborn)). Using the LOOIC goodness of fit criteria the best fitting model was found to be a Negative Binomial model that adjusted for the change in policy, age, and incidence rates in the UK born (Table \@ref(tab:07-universal-tab)). In this model there was some evidence of an assocation between the change in policy and an increase in incidence rates in those at school-age who were UK born, with an IRR of  `r extract_inc_ef(ModelComUniversalUKBorn, ModelComUniversalUKBorn$Model[1])`. Dropping the change in policy from the model resulted in a small decrease in the LOOIC (`r BestFitUniversalUKBorn[[5]]$pretty_diff_looic`) but the change was too small, with too large a standard error, to conclusively state that the excluding the change in policy from the model improved the quality of model fit. I found that it was important to adjust for UK born incidence rates, otherwise the impact from the change in BCG vaccination policy was over-estimated.

For the comparable non-UK born cohort who were relevant to the universal vaccination there was evidence, in the best fitting model, that ending the scheme was associated with a decrease in incidence rates (IRR: `r extract_inc_ef(ModelComUniversalNonUKBorn, ModelComUniversalNonUKBorn$Model[1])`). The best fitting model was a Negative Binomial model which adjusted for the change in policy, age, incidence rates in the non-UK born, and year of eligibility as a random effect (Table \@ref(tab:07-universal-tab)). I found that omitting the change in policy from the model resulted in poorer model fit (LOOIC increase of `r str_replace(BestFitUniversalNonUKBorn[[5]]$pretty_diff_looic, "-", "")`), suggesting that the policy change was an important factor explaining changes in incidence rates, after adjusting for other covariates. All models that adjusted for incidence rates in the UK born or non-UK born estimated similar IRRs (Table \@ref(tab:07-summary-universal-nonukborn)).

@@

For the UK born cohort relevant to the targeted neonatal vaccination programme the evidence of an association between the change in policy and TB incidence was mixed across all models and credible intervals were wide compared to models for the UK born cohort relevant to the universal school-age vaccination programme (Table \@ref(tab:07-summary-targeted-ukborn)). The best fitting model was a Poisson model which adjusted for the change in policy, age, UK born incidence rates, and year of study entry with a random effect (Table \@ref(tab:07-targeted-tab)). In this model, there was weak evidence of an association between the change in BCG policy and an decrease in incidence rates in UK born neonates, with an IRR of `r extract_inc_ef(ModelComTargetedUKBorn, ModelComTargetedUKBorn$Model[1])`. There was weak evidence to suggest that dropping the change in policy from this model improved the quality of the fit, with an improvement in the LOOIC score of `r BestFitTargetedUKBorn[[5]]$pretty_diff_looic`. This suggests that the change in policy was not an important factor for explaining incidence rates, after adjusting for covariates. Models which also adjusted for non-UK born incidence rates estimated that the change in policy was associated with no change in incidence rates in the relevant cohort of neonates.

For the comparable non-UK born cohort who were relevant to the targeted neonatal vaccination programme there was evidence, across all models, that the change in policy was associated with a large decrease in incidence rates (IRR: `r extract_inc_ef(ModelComTargetedNonUKBorn, ModelComTargetedNonUKBorn$Model[1])`) (Table \@ref(tab:07-targeted-tab) in the best fitting model). The best fitting model was a Negative Binomial model that adjusted for the change in policy, age, and non-UK born incidence rates (Table \@ref(tab:07-targeted-tab)). All models which at least adjusted for age estimated comparable effects of the change in policy (Table \@ref(tab:07-summary-targeted-nonukborn)).

@@

It is well established that interventions against infectious diseases, such as TB, should be evaluated not only for their direct effects but also for future indirect effects via ongoing transmission. Statistical approaches such as those used in this paper are not appropriate for capturing these future indirect effects, and instead dynamic disease models should be used. In Chapter \@ref(model-development) I develop such a dynamic disease model, Chapter \@ref(model-fitting) then fits this model to the available data, and Chapter \@ref(model-results) compares the impact of continuing with the BCG school's scheme post 2005 to universal neonatal vaccination. In addition, this study could not evaluate the impact of the neonatal programme on the high-risk population it targets, due to a lack of reliable data. Improved coverage data for the BCG programme is required to more fully evaluate its ongoing impact.

>>>>

## Read formated results
## See https://github.com/seabbs/DirectEffBCGPolicyChange/blob/master/vignettes/paper.Rmd
## for code required to reproduce 
## Chapter is an edited version of the orginal reproducible document.
##Wordcount: 4498
read_formated <- function(name) {
  path <- file.path("chapters/direct-eff-bcg-change", paste0(name, ".rds"))

@@

I combined notification data from the Enhanced TB Surveillance (ETS) system, with demographic data from the Labour Force Survey (LFS) to construct retrospective cohorts of individuals in England relevant to both the universal, and targeted vaccination programmes between Jan 1, 2000 and Dec 31, 2010. For each cohort, I estimated incidence rates over a 5 year follow-up period and used Poisson and negative binomial regression models in order to estimate the impact of the change in policy on TB. This work was adapted from a preprint^[Preprint: https://doi.org/10.1101/567511] supervised by Hannah Christensen and Ellen Brooks-Pollock. Nicky Welton provided guidance on the statistical methods used.

@@

### Constructing Retrospective cohorts {#cohort-definition}

I constructed retrospective cohorts of TB cases and individuals using the ETS and the LFS. Tuberculosis cases were extracted from the ETS based on date of birth and date of TB notification.

Cohort 1: individuals aged 14 years between 2000 and 2004, who were notified with TB while aged between 14 and 19 years.

Comparison cohort 1: individuals aged 14 years between 2005 and 2010, who were notified with TB while aged between 14 and 19 years.

Cohort 2: individuals born between 2005 and 2010, who were notified with TB while aged 0 to 5 years.

Comparison cohort 2: individuals born between 2000 and 2004, who were notified with TB while aged 0 to 5 years.

Cohorts were stratified by vaccination programme using age criteria and then stratified further by whether the scheme was in place during the time period they entered the study. Each cohort was further stratified by UK birth status, with both non-UK born and UK born cases assumed to have been exposed to England’s vaccination policy. Corresponding population cohorts were calculated using the LFS population estimates, resulting in eight population level cohorts, each with 5 years of follow-up (Table \@ref(tab:07-cohorts)).

@@

I calculated Incidence Rate Ratios (IRRs) for the change in incidence rates associated with the change in BCG vaccination policy (modelled as a binary breakpoint at the start of 2005) for both the UK born and non-UK born populations that were relevant to the universal programme, and for the targeted programme using a series of increasingly complex models. I considered the following covariates: age,[@PHE2016a; @Zwerling2011] incidence rates in both the UK born and non-UK born who were not in the age group of interest,[@PHE2016a] and year of study entry (as a random intercept). I first investigated a univariable Poisson model, followed by combinations of covariates (Table \@ref(tab:07-model-spec)). I also investigated a negative binomial model adjusting for the same covariates as in the best fitting Poisson model. The models were estimated with a Bayesian approach using Markov Chain Monte Carlo (MCMC), with default weakly informative priors (see  Section \@ref(prior-choice)). Model fit, penalised by model complexity, was assessed using the leave one out cross validation information criterion (LOOIC) and its standard error.[@Vehtari2016] Models were ranked by goodness of fit, using their LOOIC, with a smaller LOOIC indicating a better fit to the data after adjusting for the complexity of the model. No formal threshold for a change in the LOOIC was used, with changes in the LOOIC being evaluated in the context of their standard error. The inclusion of the change in policy in the best fitting model was tested by refitting the model excluding the change in policy and estimating the improvment in the LOOIC. Once the best fitting model had been identified I estimated the number of cases prevented, from 2005 until 2015, for each vaccination programme in the study population relevant to that programme (see Section \@ref(magnitude-estimation)).

@@

R `3.5.0` was used for all analysis.[@R] Reproducibility was ensured by using R package infrastructure^[Code: https://github.com/seabbs/DirectEffBCGPolicyChange]. Missing data imputation using a gradient boosting model (GBM) was implemented using the `h2o` package (see Section \@ref(imp-uk-birth-status)).[@h2o2018] Incidence rates, with 95% confidence intervals, were calculated using the `epiR` package (see Chapter \@ref(data)).[@EpiR] The `brms` package,[@Burkner] and STAN,[@StanDevelopmentTeam2016] was used to perform Markov Chain Monte Carlo (MCMC). Models were run until convergence (4 chains with a burn in of 10,000, and 10,000 sampled iterations each), with convergence being assessed using trace plots and the R hat diagnostic.[@StanDevelopmentTeam2016] All numeric confounders were centered and scaled by their standard deviation, and age was adjusted for using single year of age categories.

@@

As I was imputing a single variable, I reformulated the imputation as a categorical prediction problem. This allowed the use of more complex, high-performing models compared to those usually used for imputation, whilst also allowing the results to be validated using predictive modelling performance metrics. I included year of notification, sex, age, PHE Centre (PHEC), occupation, ethnic group, Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank), and risk factor count (risk factors considered; drug use, homelessness, alcohol misuse/abuse and prison). However, I could not account for a possible missing not at random mechanism not captured by these covariates. To train the model I first split the data with complete UK birth status into a training set (80%), a calibration set (5%), and a test set (15%). I then fit a gradient boosted machine with 10,000 trees, early stopping (at a precision of $1e-5$, with 10 stopping rounds), a learning rate of 0.1, and a learn rate annealing of 0.99. Gradient boosted machines are a tree based method that can incorporate complex non-linear relationships and interactions. Much like a random forest model they work by ensembling a group of trees, but unlike a random forest model each tree is additive aiming to reduce the residual loss from previous trees. Once the model had been fit to the training set I performed platt scaling (fits a logistic regression model to model predictions in order to return a probability) using the calibration dataset. The fitted imputation model had a logloss (the negative of the log likelihood) of 0.28 on the test set, with an area under the curve (AUC) of 0.93, both of which indicate robust performance on unseen data. I found that ethnic group was the most important variable for predicting UK birth status, followed by age and PHEC.

@@

I estimated the magnitude of the estimated impact from the change in BCG policy by applying the IRR estimates from the best fitting model for each cohort to the observed number of notifications from 2005 until 2015 in the study population. For the cohorts relevant to the universal school-age vaccination scheme I estimated the number of prevented cases by first aggregating cases ($C_0$) and then using the following equation,

\[ C^i_P = C_0 * (1 - I^i),\ \text{Where}\ i = e,\ l,\ u.\]

@@

In the UK born cohort relevant to universal vaccination there was some evidence, across all models that adjusted for age, that ending the scheme was associated with a modest increase in TB rates (Table \@ref(tab:07-summary-universal-ukborn)). Using the LOOIC goodness of fit criteria the best fitting model was found to be a negative binomial model that adjusted for the change in policy, age, and incidence rates in the UK born (Table \@ref(tab:07-universal-tab)). In this model there was some evidence of an assocation between the change in policy and an increase in incidence rates in those at school-age who were UK born, with an IRR of  `r extract_inc_ef(ModelComUniversalUKBorn, ModelComUniversalUKBorn$Model[1])`. Dropping the change in policy from the model resulted in a small decrease in the LOOIC (`r BestFitUniversalUKBorn[[5]]$pretty_diff_looic`) but the change was too small, with too large a standard error, to conclusively state that the excluding the change in policy from the model improved the quality of model fit. I found that it was important to adjust for UK born incidence rates, otherwise the impact from the change in BCG vaccination policy was over-estimated.

For the comparable non-UK born cohort who were relevant to the universal vaccination there was evidence, in the best fitting model, that ending the scheme was associated with a decrease in incidence rates (IRR: `r extract_inc_ef(ModelComUniversalNonUKBorn, ModelComUniversalNonUKBorn$Model[1])`). The best fitting model was a negative binomial model which adjusted for the change in policy, age, incidence rates in the non-UK born, and year of eligibility as a random effect (Table \@ref(tab:07-universal-tab)). I found that omitting the change in policy from the model resulted in poorer model fit (LOOIC increase of `r str_replace(BestFitUniversalNonUKBorn[[5]]$pretty_diff_looic, "-", "")`), suggesting that the policy change was an important factor explaining changes in incidence rates, after adjusting for other covariates. All models that adjusted for incidence rates in the UK born or non-UK born estimated similar IRRs (Table \@ref(tab:07-summary-universal-nonukborn)).

@@

For the UK born cohort relevant to the targeted neonatal vaccination programme (see Section \@ref(cohort-definition)) the evidence of an association between the change in policy and TB incidence was mixed across all models and credible intervals were wide compared to models for the UK born cohort relevant to the universal school-age vaccination programme (Table \@ref(tab:07-summary-targeted-ukborn)). The best fitting model was a Poisson model which adjusted for the change in policy, age, UK born incidence rates, and year of study entry with a random effect (Table \@ref(tab:07-targeted-tab)). In this model, there was weak evidence of an association between the change in BCG policy and an decrease in incidence rates in UK born neonates, with an IRR of `r extract_inc_ef(ModelComTargetedUKBorn, ModelComTargetedUKBorn$Model[1])`. There was weak evidence to suggest that dropping the change in policy from this model improved the quality of the fit, with an improvement in the LOOIC score of `r BestFitTargetedUKBorn[[5]]$pretty_diff_looic`. This suggests that the change in policy was not an important factor for explaining incidence rates, after adjusting for covariates. Models which also adjusted for non-UK born incidence rates estimated that the change in policy was associated with no change in incidence rates in the relevant cohort of neonates.

For the comparable non-UK born cohort who were relevant to the targeted neonatal vaccination programme there was evidence, across all models, that the change in policy was associated with a large decrease in incidence rates (IRR: `r extract_inc_ef(ModelComTargetedNonUKBorn, ModelComTargetedNonUKBorn$Model[1])`) (Table \@ref(tab:07-targeted-tab) in the best fitting model). The best fitting model was a negative binomial model that adjusted for the change in policy, age, and non-UK born incidence rates (Table \@ref(tab:07-targeted-tab)). All models which at least adjusted for age estimated comparable effects of the change in policy (Table \@ref(tab:07-summary-targeted-nonukborn)).

@@

It is well established that interventions against infectious diseases, such as TB, should be evaluated not only for their direct effects but also for future indirect effects via ongoing transmission. Statistical approaches such as those used in this chapter are not appropriate for capturing these future indirect effects, and instead dynamic disease models should be used. In Chapter \@ref(model-development) I develop such a dynamic disease model, Chapter \@ref(model-fitting) then fits this model to the available data, and Chapter \@ref(model-results) compares the impact of continuing with the BCG school's scheme post 2005 to universal neonatal vaccination. In addition, this study could not evaluate the impact of the neonatal programme on the high-risk population it targets, due to a lack of reliable data. Improved coverage data for the BCG programme is required to more fully evaluate its ongoing impact.
